**Effective application of machine learning to a microbiome-based classification problem**

### 2020-01-17 submission/manuscript.Rmd
some additions for rebuttal

---
title: '**Effective application of machine learning to microbiome-based classification problems**'
csl: mbio.csl
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
bibliography: references.bib
---

@@

Running title: Machine learning methods applied to microbiome studies

@@

Prediction of health outcomes using machine learning (ML) is rapidly being adopted in microbiome studies. However, the estimated performance associated with these ML models is likely over-optimistic. Moreover, there is a trend towards using black box models without a discussion of the difficulty of interpreting such models when trying to identify microbial biomarkers of disease. This work represents a step towards developing more reproducible ML practices in applying ML to microbiome research. We implement a rigorous pipeline and emphasize the importance of selecting ML models that reflect the goal of the study. These concepts are not particular to the study of health outcomes but can also be applied to environmental microbiology studies.

@@

Among microbiome researchers, the lack of justification when selecting a modeling approach has often been due to an implicit assumption that more complex models are better. This has resulted in a trend towards using non-linear models such as random forest and deep neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019]. Although in some cases, complex models may capture important non-linear relationships and therefore yield better predictions, they can also result in black boxes that lack interpretability. Such models require post hoc explanations to quantify the importance of each feature in making predictions. Depending on the goal of the modeling, other approaches may be more appropriate. For example, researchers trying to identify the microbiota associated with disease may desire a more interpretable model, whereas clinicians may emphasize predictive performance. Nonetheless, it is essential to understand that the benefit of more complex, less interpretable models may be minimal [@rudin_please_2018; @rudin_optimized_2018]. It is important for researchers to justify their choice of modeling approach.

To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing seven modeling approaches with the same dataset and pipeline. We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel. We also trained four non-linear models: SVM with radial basis function kernel, a decision tree, random forest, and gradient boosted trees. We compared their predictive performance, interpretability, and training time. To demonstrate the performance of these modeling approaches and our pipeline, we used data from a previously published study that sought to classify individuals as having healthy colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016]. This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011]. With this dataset, we developed an ML pipeline that can be used in many different scenarios for training and evaluating models. This framework can be easily applied to other host-associated and environmental microbiome datasets.

@@

__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having healthy colons or SRNs [Figure 2]. The predictive performance of random forest model was higher than other ML models with a median  `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`], though not significantly (p=0.5) (Figure S3). Similarly, L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(median_iqr(all, "L2_Logistic_Regression")[1])` [IQR `r paste(median_iqr(all, "L2_Logistic_Regression")[2])`-`r paste(median_iqr(all, "L2_Logistic_Regression")[3])`], `r paste(median_iqr(all, "XGBoost")[1])` [IQR `r paste(median_iqr(all, "XGBoost")[2])`-`r paste(median_iqr(all, "XGBoost")[3])`], `r paste(median_iqr(all, "L2_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L2_Linear_SVM")[2])`-`r paste(median_iqr(all, "L2_Linear_SVM")[3])`] and `r paste(median_iqr(all, "RBF_SVM")[1])` [IQR `r paste(median_iqr(all, "RBF_SVM")[2])`-`r paste(median_iqr(all, "RBF_SVM")[3])`], respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(median_iqr(all, "L1_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L1_Linear_SVM")[2])`-`r paste(median_iqr(all, "L1_Linear_SVM")[3])`] and `r paste(median_iqr(all, "Decision_Tree")[1])` [IQR `r paste(median_iqr(all, "Decision_Tree")[2])`-`r paste(median_iqr(all, "Decision_Tree")[3])`], respectively [Figure 2]. Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and L2-regularized SVM with linear kernel) performed nearly as well as non-linear models.

@@

**Figure S3. Histogram of AUROC differences between L2-regularized logistic regression and random forest for each of the hundred data-splits.** This histogram shows the number of data-splits in each bin. The percentage of dataplits where the difference between random forest and L2-regularized logistic regression AUROC values was higher than or equal to 0 were 0.75, lower than or equal to 0 were 0.25. The vertical red line highlights the bins where there AUROC difference between the two model is 0. The p-value was calculated for a double tail event. 
\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}

>>>>

---
title: '**Effective application of machine learning to a microbiome-based classification problem**'
csl: mbio.csl
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
bibliography: references.bib
---

@@

Running title: Machine learning methods applied to a microbiome study

@@

Prediction of diseases using machine learning (ML) is rapidly being adopted in microbiome studies. However, the estimated performance associated with these ML models is likely over-optimistic. Moreover, there is a trend towards using black box models without a discussion of the difficulty of interpreting such models when trying to identify microbial biomarkers of disease. This work represents a step towards developing more reproducible ML practices in applying ML to microbiome research. We implement a rigorous pipeline and emphasize the importance of selecting ML models that reflect the goal of the study. These concepts are not particular to the study of human health but can also be applied to environmental microbiology studies.

@@

Among microbiome researchers, the lack of justification when selecting a modeling approach has often been due to an implicit assumption that more complex models are better. This has resulted in a trend towards using non-linear models such as random forest and deep neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019]. Although in some cases, complex models may capture important non-linear relationships and therefore yield better predictions, they can also result in black boxes that lack interpretability. Such models require post hoc explanations to quantify the importance of each feature in making predictions. Depending on the goal of the modeling, other approaches may be more appropriate. For example, researchers trying to identify the microbiota associated with disease may desire a more interpretable model, whereas clinicians may emphasize predictive performance. Nonetheless, it is essential to understand that the benefit of more complex, less interpretable models may be minimal [@rudin_please_2018; @rudin_optimized_2018; @Quinn847475]. It is important for researchers to justify their choice of modeling approach.

In this study, we provided steps toward standardization of machine learning methods for microbiome studies which are often poorly documented and executed. To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing seven modeling approaches with the same dataset and pipeline. We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel. We also trained four non-linear models: SVM with radial basis function kernel, a decision tree, random forest, and gradient boosted trees. We compared their predictive performance, interpretability, and training time. To demonstrate the performance of these modeling approaches and our pipeline on a case study, we used a data from a previously published study that sought to classify individuals as having healthy colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016]. This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011]. With this dataset, we developed an ML pipeline that can be used in many different scenarios for training and evaluating models. This framework can be easily applied to other host-associated and environmental microbiome datasets. We hope that this study ...

@@

__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having healthy colons or SRNs [Figure 2]. The predictive performance of random forest model was higher than other ML models with a median  `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`], though not significantly (p=0.5; The p-value was manually calculated using the sampling distribution of the test statistic under the null hypothesis) (Figure S3). Similarly, L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(median_iqr(all, "L2_Logistic_Regression")[1])` [IQR `r paste(median_iqr(all, "L2_Logistic_Regression")[2])`-`r paste(median_iqr(all, "L2_Logistic_Regression")[3])`], `r paste(median_iqr(all, "XGBoost")[1])` [IQR `r paste(median_iqr(all, "XGBoost")[2])`-`r paste(median_iqr(all, "XGBoost")[3])`], `r paste(median_iqr(all, "L2_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L2_Linear_SVM")[2])`-`r paste(median_iqr(all, "L2_Linear_SVM")[3])`] and `r paste(median_iqr(all, "RBF_SVM")[1])` [IQR `r paste(median_iqr(all, "RBF_SVM")[2])`-`r paste(median_iqr(all, "RBF_SVM")[3])`], respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(median_iqr(all, "L1_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L1_Linear_SVM")[2])`-`r paste(median_iqr(all, "L1_Linear_SVM")[3])`] and `r paste(median_iqr(all, "Decision_Tree")[1])` [IQR `r paste(median_iqr(all, "Decision_Tree")[2])`-`r paste(median_iqr(all, "Decision_Tree")[3])`], respectively [Figure 2]. Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and L2-regularized SVM with linear kernel) performed nearly as well as non-linear models.

@@

**Figure S3. Histogram of AUROC differences between L2-regularized logistic regression and random forest for each of the hundred data-splits.** This histogram shows the number of data-splits in each bin. The percentage of dataplits where the difference between random forest and L2-regularized logistic regression AUROC values was higher than or equal to 0 were 0.75, lower than or equal to 0 were 0.25. The vertical red line highlights the bins where there AUROC difference between the two model is 0. The p-value was manually calculated using the sampling distribution of the test statistic under the null hypothesis. We are testing how often random forest performs more accurately than L2-regularized logistic regression. We are assuming that the distribution of the differences between the AUROC values of random forest and L2 logistic regression is symmetric about 0, therefore we are calculating the p-value for a double-tail event.
\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}
